---
title: "working_thesis_ACTIVITIES"
format: html
---

Design: Jung et al., "Style over Story" (arXiv:2510.02025v4)

# packages
```{r}
library(tidyverse)
library(readxl)
library(glue) # create activity sentence
library(sylcount) # check sentence metrics 
```


# Activity library

Structure: 15 topics x 3 actors x 2 levels = 90 activities.

- **primary hidden axis =** `actor` (upstream, midstream, downstream)

- **secondary hidden axis =** `level` (low-cognitive complexity,
  high-cognitive complexity)

- `topic` is the consumption proxy for each IPCC mitigation strategy
  (appliances; solar; refrigerants; travel; insulation; wood products;
  plastics; smartphones; food waste; heating; urban greening; clothing;
  steel)

## Load action bank

```{r}
actions <- read_excel("data/activity_bank.xlsx") 
actions
```

## Wrap each action in the sentence template

Every activity is the same sentence template with one action inserted
into it.

```{r}
sentence_template <- "Students examine recommendations to {action} and determine how the strategy could limit anthropogenic impact on Earth's systems."

activity <- actions |>
  mutate(activity = as.character(
    glue_data(pick(everything()), 
              sentence_template)))

activity 

```

## Build the library

Build text to send to LLM and save to csv.

```{r}
# Generate random activity ID 
# OPTION 
pool <- as.vector(outer(
  as.vector(outer(LETTERS, LETTERS, paste0)),
  sprintf("%02d", 0:99), paste0))
lib <- activity |>
  arrange(topic, actor, level) |>
  mutate(activity_id = paste0("a_", sample(pool, n())))

set.seed(123)
activity_list <- activity |>
  arrange(topic, actor, level) |> 
  # add ID to match LLM selection 
  mutate(activity_id = replicate(n(),
         paste(c("a_",
                 sample(LETTERS, 2, replace = TRUE),
                 sample(0:9, 2, replace = TRUE)), 
                      collapse = "")))
  
write_csv(activity_list, "data/activity_main.csv")

```

# Sentence metrics

Items were held constant in surface form. Length ranged from 24 to 28 words (M = 25.7) and syllables per word from 1.75 to 2.24 (M = 2.03).

```{r}
# Check that sentence template and unique IDs worked 
n_distinct(activity_list$activity)
n_distinct(activity_list$activity_id)


# Metrics  
# REDO THIS FOR OTHER METRICS SUITABLE FOR SENTENCE 
read_stats <- readability(activity_list$activity) |>
  as_tibble() |>
  select(n_chars = chars, # Characters per sentence
         n_words_total = words, # Total words per sentence 
         n_sents = sents, # Check how many sentences
         n_syll = sylls) # number of syllables in the sentence

metrics <- activity_list |>
  bind_cols(read_stats) |>
  mutate(
    syll_per_word = n_syll / n_words_total,
    mean_word_len = n_chars / n_words_total)

metrics


# Overall
metrics |>
  summarise(min = min(n_words_total), 
            max = max(n_words_total),
            mean = round(mean(n_words_total), 1),
            sd = round(sd(n_words_total), 2))

# By cell 
metrics |>
  group_by(actor, level) |>
  summarise(n = n(),
            words_min = min(n_words_total),
            words_max = max(n_words_total),
            words_mean = round(mean(n_words_total), 1),
            words_sd = round(sd(n_words_total), 2),, 
            syl_min = min(syll_per_word), 
            syl_max = max(syll_per_word),
            .groups = "drop") |>
  mutate(words_range = words_max - words_min)

```


