---
title: "API_data"
format: html
---

# Design

Replication for Jung et al., "Style over Story" (arXiv:2510.02025v4).

Three LLMs are each shown the same list of 90 learning activities and asked to select the ones they would teach. The activities vary on two hidden axes (`actor` and `level`) that are never shown to the model. What we want to know is whether the system prompt (`basic` / `quality` / `creativity`) shifts *which* axis the model favours.


## Packages

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(tibble)
library(readr)
library(readxl)
library(glue)
library(httr2)
library(jsonlite)
library(withr)
```

##Data - load/save
Save what is sent and all responses - same folders as Stage 1 but different file names. 

```{r}
paths <- list(
  response = "response",
  sent = "sent",
  
# Save list order for each run
  presented = file.path("sent", "list_order_s1.csv"),

# Main run files: 
# metadata: one row per run (model, instruction, condition, seed, status and tokens)
  runs = file.path("response", "metadata.csv"),
# raw JSON responses
  raw = file.path("response", "raw_responses.rds"),
# tidied responses
  parsed = file.path("response", "selections_long.csv"), 

# Pilot files:
  runs_pilot = file.path("response",
                         "metadata_pilot.csv"),
  raw_pilot = file.path("response", 
                        "raw_responses_pilot.rds"),
  parsed_pilot = file.path("response",
                           "selections_long_pilot.csv"), 
# Retry files:
  runs_retry = file.path("response",
                         "metadata_retry.csv"),
  raw_retry = file.path("response",
                         "raw_responses_retry.rds"))

# Create folders  
#################################################
# Does not need to be run if API_Data_s1 has already run but will not overwrite anything if it already exists. 
dir.create(paths$response,
           showWarnings = FALSE, 
           recursive = TRUE)
dir.create(paths$sent,
           showWarnings = FALSE, 
           recursive = TRUE)

# Load activities list
activities <- read_csv("data/activity_main.csv", 
                show_col_types = FALSE)
activities

```

## Settings

- `actor_names` and `level_names` are the two hidden axes.

- `k_fixed` is the budget in the fixed conditions.

- `m_seed` seeds the presentation order so a run can be reproduced.

```{r}
# set k_fixed
k_fixed <- 15

# Max number of tokens models can generate in their response 
# Note: higher = more costs but run will fail if limit is met 
# sonnet5 - adaptive thinking requires more tokens 
max_tokens <- 20000

# max number of API requests if initially failed
# Prevents infinite loops 
max_tries <- 5

# pause for 1 second before / after next request 
req_pause <- 1.0

# Global seed for reproducibility
m_seed <- 2026
set.seed(m_seed)

################################################################3
## FIX THIS UP 
# Run for cleaning text when matching 
normalise_text <- function(x) {
  x |> 
    str_replace_all("[^A-Za-z0-9 ]", " ") |> 
    str_to_lower() |> 
    str_squish()
}
# TO REPLACE NORM AND KEY 
# normalise and clean text string in case of minor changes 
# removes random spaces, makes everything lowercase
norm <- \(x) str_squish(tolower(
  # remove puntuation, new lines etc. 
  str_replace_all(x,"[^A-Za-z0-9 ]", " ")))
```

#Content to send 

Includes the construction and formatting for everything that is sent to the models.

## System prompts

Three instruction types to address response sensitivity caused by wording of the prompt.
These are the roles sent to each model (Anthropic - `system`; OpenAI - `instructions`; Google - `system_instructions`)

```{r}
instruction_types <- tribble(
  ~instruction_type, ~system_prompt,

  "basic",
  "You are an Australian secondary educator. Your task is to select activities when requested. Your goal is to select activities that fulfill the given requirements.",

  "quality",
  "You are a highly skilled Australian secondary educator known for technical excellence and flawless execution of pedagogical fundamentals. You design activities with precise task sequencing, well-structured lessons, clear curriculum alignment and carefully planned differentiation. Your goal is to design activities of the highest quality through careful refinement and technical mastery.",

  "creativity",
  "You are an innovative Australian secondary educator celebrated for creating completely original and unexpected learning activities. You excel at breaking conventional lesson structures and exploring new pedagogical possibilities. Your strength lies in developing unusual task sequences, unconventional lesson structures, or experimental modes of student participation that students would not expect. Your goal is to design activities that are unlike anything that has been taught before, pushing the boundaries of what learning can be through creative experimentation."
)

instruction_types
```

## User prompt

Two experiment conditions. Prompt 1 (K ≥ 0) and Prompt 2 (K = 15).
This is the experiment condition sent to the models (Anthropic - `messages`; OpenAI - `input`; Google - `contents`)

```{r}

prompt_1 <- r"(As you plan to teach a unit, identify the specific activities that would be most useful for a single 10-week unit on Earth systems and climate change, and explain your reasoning for why each activity would help support better student learning.  

Task:
-	You will be given a list of 90 possible learning activities.
-	Read through all 90 activities carefully.
-	Select the activities you consider most useful for teaching a unit on Earth systems and climate change.
-	You may select as few or as many activities as you believe are appropriate. There is no minimum or maximum.
-	For each selected activity, explain your reason for choosing it.
-	There is a limit of one or two sentences for each of your explanations. There are no restrictions on the style of your explanations. 
-	You do not need to mention activities you are not selecting unless you wish to explain why you excluded them.
-	List your selections using the specified output format for easy parsing.

Output Format:
-	You may select as few or as many activities as you wish. The order in which you list them does not matter.
-	For each, write only the selected activity as a JSON object, then the activity ID in the "activity_id" field and your reason in the "reason" field.
-	Each activity, activity ID and its reason must appear as a separate element in a single JSON array containing all elements.
- Respond with the JSON array only. Do not write anything before or after it. 

Example Output: {example_output}
Activities List: {activity_list})"

prompt_2 <- r"(As you plan to teach a unit, identify the specific activities that would be most useful for a single 10-week unit on Earth systems and climate change, and explain your reasoning for why each activity would help support better student learning.
Task:
-	You will be given a list of 90 possible learning activities.
-	Read through all 90 activities carefully.
-	Select exactly 15 activities you consider most useful for teaching a unit on Earth systems and climate change.
-	For each selected activity, explain your reason for choosing it.
-	There is a limit of one sentence for each of your explanations. There are no restrictions on the style of your explanations. 
-	You do not need to mention activities you are not selecting unless you wish to explain why you excluded them.
-	List your selections using the specified output format for easy parsing.

Output Format:
-	Select exactly 15 activities. The order in which you list them does not matter.
-	For each, write only the selected activity as a JSON object, then the activity ID in the "activity_id" field and your reason in the "reason" field.
-	Each activity, activity ID and its reason must appear as a separate element in a single JSON array containing all elements.
- Respond with the JSON array only. Do not write anything before or after it. 

Example Output: {example_output}
Activities List: {activity_list})"

```

## Example output

A worked example of the required JSON shape to include with both prompts.

```{r}
example_output <- paste0('[\n','{"activity": "Students examine safety features in cars and determine how each feature relates to the Laws of Motion.", "activity_id": "b_XF59", "reason": "<one or two sentences>"},\n',
  '{"activity": "Students examine how elements are organised on the periodic table and determine patterns that group elements together.", "activity_id": "b_GK63", "reason": "<one or two sentences>"}\n',']')
# check 
cat(example_output)
```

##Activity list 

```{r}
# Ensure no other information is sent to LLM so the model only ever sees the activity and the ID
activities_send <- activities |> 
  select(activity_id, activity) 
```

##Render activity list

The order of the activity list is shuffled for every run to address selection preference sensitivity caused by the order or position in which the activities are presented. When the grid is established, each run is assigned a unique `presentation_seed` (`m_seed` + run number) used for `set.seed(seed)` inside `render_list()`. This ensures that the order of each list rendered by `sample.int()` is only dependent on its distinct seed per run, making each permutation recordable prior to running the API loop and also repeatable for reattempting any failed runs. 

```{r}

# Function for randomising list order 
render_list <- function(activity_shuf, seed) {
  set.seed(seed)
    # Shuffle the list for one run and number each row
    activity_shuf |>
      # whole row (so ID and activity are constant)
      slice(sample.int(nrow(activity_shuf))) |>
      # Position to save list order 
      mutate(position = row_number()) |>
      # only rows available but in case of future changes
      select(position, activity_id, activity)
}

# function - shuffled tibble into plain text block to paste into the prompt 
format_list_for_prompt <- function(send) {
  # one line for ID and activity then new line. 
  paste("ID:",
         send$activity_id, 
         "| Activity:",
         send$activity,
         collapse = "\n")}

# Check sample
send_demo <- render_list(activities_send, 
                         seed = m_seed)
# Shorten to 500 characters
cat(str_trunc(format_list_for_prompt(send_demo), 500))

# Check activity shuffle never drops or changes an activity and that the random order is working
# seed = m_seed + row number (e.g., run 1 seed is 2027)
order_1 <- render_list(activities_send, seed = 2027)
order_1a <- render_list(activities_send, seed = 2027)
order_2 <- render_list(activities_send, seed = 2028) 
# (should all be TRUE)
    # different seed = different order
cat(!identical(order_1, order_2),
    # same seed = same order 
    identical(order_1, order_1a),
    # order_1 IDs same as original IDs (regardless of order)
    setequal(order_1$activity_id,
             activities_send$activity_id))

```

## Conditions

```{r}
# add condition label and budget to the prompt templates
conditions <- tribble(
  ~condition, ~k, ~template,
  "k_free",  NA_real_, prompt_1,
  "k_fixed", k_fixed, prompt_2)

build_user_prompt <- function(condition_row, 
                              activity_list) {
  # fill {} from the prompt 
  glue(condition_row$template, 
       .trim = FALSE, 
       .envir = environment()) |> 
    as.character()}

conditions
```


##User prompt 
User prompt includes the condition instructions, activity list and example output. 

```{r}

build_user_prompt <- function(condition_row, 
                              activity_list) {
  # fills the prompt templates placeholders 
  glue(condition_row$template, 
       # preserves the prompts structure 
       .trim = FALSE, 
# sets the environment to inside the function call for {activity_list} 
# moves to global environment for {example_output}
       .envir = environment()) |> 
    # character vector
    as.character()}

```


##Check one built prompt before sending.

```{r}
# check that one fully rendered prompt
demo_prompt <- build_user_prompt(
  condition_row = conditions[conditions$condition == "k_fixed", ],
  activity_list = format_list_for_prompt(send_demo))

cat(demo_prompt)
```

# Model APIs

**Open AI** 
* GPT-5.6 Terra
Release Date: 9th July, 2026.
Knowledge and training data cutoff: 16th February, 2026.
https://developers.openai.com/api/docs/models/gpt-5.6-terra 
https://developers.openai.com/api/reference/resources/responses 

**Anthropic** 
* Sonnet 5
Release Date: 30th June, 2026. 
Knowledge and training data cutoff: January 2026. 
https://platform.claude.com/docs/en/models/sonnet-5/overview 
https://platform.claude.com/docs/en/api/http/messages/create 

**Gemini** 
* Gemini 3.8 Flash
Release Date: September, 2026.
Knowledge and training data cutoff: March, 2026. 
https://ai.google.dev/gemini-api/docs/models/gemini-3.8-flash 
https://ai.google.dev/gemini-api/docs 

NOTE: new models do not support top_p or temperature params

## Models and Default Settings 

```{r}
# create model vars and parameters (default settings)
models <- tribble(
  
# one row per model; params is a list column with provider settings
  ~model, ~provider, ~model_id, ~params,

  "gpt_terra", "openai", "gpt-5.6-terra", 
  list(reasoning = list(effort = "medium"), store = FALSE), 
  # reasoning default effort = medium 
  
  "sonnet5", "anthropic", "claude-sonnet-5",
  list(output_config = list(effort = "high")), 
  # adaptive reasoning default effort = high 
  
  "gemini3_8", "google", "gemini-3.8-flash",
  list(thinkingConfig = list(thinkingLevel = "medium")))
  # reasoning default effort = medium 

models
```


## Endpoints

```{r}
# Set endpoints
endpoints <- list(
  anthropic = "https://api.anthropic.com/v1/messages",
  openai = "https://api.openai.com/v1/responses",
  #Google embeds model id in URL so %s placeholder is used incase models change - filled by sprintf()
  google = "https://generativelanguage.googleapis.com/v1beta/models/%s:generateContent")

# set keys 
api_keys <- c(
  anthropic = "ANTHROPIC_API_KEY",
  openai = "OPENAI_API_KEY",
  google = "GEMINI_API_KEY")

```

##API Keys
Keys live in `~/.Renviron`. Add setup to README: run ``, add the three lines and restart R.

```{r}
#| eval: false
usethis::edit_r_environ()
# ANTHROPIC_API_KEY={your_key}
# OPENAI_API_KEY={your_key}
# GEMINI_API_KEY={your_key}
```


## Check that keys are valid 
```{r}
# check all keys are set and valid - function incase models are added in future
check_keys <- function(providers = unique(models$provider)) {
  out <- tibble()

  for (p in providers) {
    out <- bind_rows(out, tibble(
      provider = p,
      env_var = api_keys[[p]],
      # keys valid - missing or empty 
      set = nzchar(Sys.getenv(api_keys[[p]], 
                              unset = ""))))
  }
  out
}
check_keys()
```

## Function for base request 

```{r}
# similar to above - used for request to fail with message rather than running for all requests. 
get_api_key <- function(provider) {
  # Check if a model is added but not API key
  if (!provider %in% names(api_keys)) stop("Unknown provider: ",
                                           provider)
  var <- api_keys[[provider]]
  # Check API key is valid
  key <- Sys.getenv(var, unset = "")
  if (!nzchar(key)) {
    stop(var, 
         " is not set. Run usethis::edit_r_environ(), add the line, then restart R.")}
  key
}

```

## Building a request

`request()` builds a placeholder holding the endpoint and the headers, and everything else is an add-on. These are `POST` endpoints, so the parameters go in a body via `req_body_json` instead of in the URL via `req_url_query`.

`base_request` is the placeholder (endpoint plus authentication), no content included. 

Anthropic: " When making API requests, you must send an anthropic-version request header. Anthropic recommends using the latest API version whenever possible. Previous versions are considered deprecated and may be unavailable for new users. `2023-06-01` " REF: https://platform.claude.com/docs/en/api/versioning 

```{r}
# set endpoints and headers 
base_request <- function(provider, model_id) {
  if (provider == "anthropic") {
    request(endpoints$anthropic) |> #endpoints
      req_headers( 
        "x-api-key" = get_api_key("anthropic"), # Headers 
        "anthropic-version" = "2023-06-01",
        "content-type" = "application/json")

  } else if (provider == "openai") {
    request(endpoints$openai) |>
      req_headers(
        "Authorization" = paste("Bearer",
                                get_api_key("openai")),
        "content-type" = "application/json")

  } else if (provider == "google") {
    request(sprintf(endpoints$google, model_id)) |>
      req_headers(
        "x-goog-api-key" = get_api_key("google"),
        "content-type" = "application/json")

  } else {
    stop("Unknown provider: ", provider) 
  }
}

```



## Build Request Body 

Provider-specific part. 
System prompt is `system` / `instructions` / `system_instruction`, and the token limit is `max_tokens` / `max_output_tokens` / `maxOutputTokens`:

```{r}
# Messages normalised
build_req_body <- function(provider, model_id,
                          system_prompt, # role 
                          user_prompt, # experiment conditions 
                          params, # effort/reasoning 
                          n_tokens) {

  if (provider == "anthropic") {
    c(list(
        model = model_id,
        max_tokens = n_tokens,
        system = system_prompt,
        messages = list(list(role = "user", 
                             content = user_prompt))),
      params)

  } else if (provider == "openai") {
    c(list(
        model = model_id,
        instructions = system_prompt,
        input = user_prompt,
        max_output_tokens = n_tokens),
      params)

  # Create request body as a separate list
  } else if (provider == "google") {
          # System instructions - Role
    list(system_instruction = list(
      parts = list(list(text = system_prompt))),
          # user prompt - experiment conditions 
      contents = list(
        list(
          role = "user",
          parts = list(
            list(text = user_prompt)))),
      # Generation config - top_p etc no longer accepted 
      generationConfig = c(list(
        maxOutputTokens = n_tokens), 
        params)) # params sit inside generationConfig 
  } else {
    stop("Unknown provider: ", provider)
  }
}

```



# Final Request
Put the two together and you have a request ready to perform:

```{r}
build_request <- function(provider, 
                          model_id, 
                          system_prompt, 
                          user_prompt,
                          params = list(), 
                          n_tokens = max_tokens,
                          n_tries = max_tries) {

  body <- build_req_body(provider, 
                     model_id, 
                     system_prompt,
                     user_prompt, 
                     params, 
                     n_tokens)

  base_request(provider, model_id) |>
    req_body_json(body, auto_unbox = TRUE) |>
    req_timeout(600) |>
    req_error(body = resp_body_string) |> 
    req_retry(max_tries = n_tries)
}
```


## Run Grid 
```{r}
set.seed(m_seed)

n_reps <- 30

grid_run <- expand_grid(
  model = models$model,
  instruction_type = instruction_types$instruction_type,
  condition = conditions$condition,
  replicate = seq_len(n_reps)) |>
  left_join(conditions, by = "condition") |>
  left_join(instruction_types, by = "instruction_type") |>
  left_join(models, by = "model") |>
  mutate(
    run_id = sprintf("run_%03d", row_number()),
    # pre-determine seed to API loop smaller 
    # m_seed (2026) plus row number for different seed when ordering list
    presentation_seed = m_seed + row_number())

# Stop if NA prompts from failed join 
stopifnot(nrow(grid_run) == n_reps * nrow(conditions) *
            nrow(instruction_types) * sum(models),
          !any(is.na(grid_run$system_prompt)),
          !any(is.na(grid_run$template)),
          !any(duplicated(grid_run$run_id)))

```


## Shuffled activity list 

Save the order of the activity list sent with each run. Only need to run this chunk once to save file. 
```{r}
#| eval: false 
set.seed(m_seed)
# Save presentation order of activities across all runs based on presentation_seed
presented_orders <- grid_run |>
  select(run_id, presentation_seed) |>
  pmap_dfr(function(run_id, presentation_seed) {
    render_list(activities_send, 
                seed = presentation_seed) |>
      mutate(run_id = run_id, 
             presentation_seed = presentation_seed, 
             .before = 1)
  })

# Check activity order is the same as demo_prompt
cat(str_trunc(format_list_for_prompt(order_1), 600))

```

### Save shuffled activity list 

Save once - not used again, just if further analysis is ever done. 
```{r}
#| eval: false 
write_csv(presented_orders, paths$presented)
```

# Responses 
# Extract text and token metadata across providers 

Parse response to handle each providers JSON response  to extract raw text (inc. input and output token counts). 

```{r}
parse_api_response <- function(response, provider) {
  # check response is json
  body <- resp_body_json(response)
  # pull text from json blocks, drop any NULLs with no text field
  # paste(collapse = "") joins everything into one string and 
  pull_text <- function(x) paste(map_chr(x, "text", 
                                         .default = ""), 
                            collapse = "")
  
  json_blocks <- function(blocks, type){
    blocks[map_chr(blocks, "type",
                   .default = NA_character_) %in% type]
  }
  
  # Anthropic content is a thinking blovk with no text field - select by type not position 
  out <- if (provider == "anthropic") {
    list(text = pull_text(keep(body$content, 
                               \(b) identical(b$type, 
                                              "text"))),
         stop_reason = body$stop_reason,
         in_tok = body$usage$input_tokens,
         out_tok = body$usage$output_tokens)
  
  # OpneAI content has two nested levels - select message then content, flatten  
  } else if (provider == "openai") {
    blocks <- body$output |> 
      keep(\(b) identical(b$type, "message")) |> 
      map("content") |> list_flatten()
    list(text = pull_text(keep(blocks, 
                               \(b) identical(b$type,
                                              "output_text"))),
         stop_reason = body$incomplete_details$reason %||% body$status,
    in_tok = body$usage$input_tokens,
    out_tok = body$usage$output_tokens)
    
  } else if (provider == "google") {
    list(text = pull_text(body$candidates[[1]]$content$parts %||% list()),
         stop_reason = body$candidates[[1]]$finishReason,
         in_tok = body$usageMetadata$promptTokenCount,
         out_tok = body$usageMetadata$candidatesTokenCount)
    
  } else stop("Unknown provider: ", provider)
  
    stopifnot(all(c("text", 
                    "stop_reason", 
                    "in_tok", 
                    "out_tok") %in% names(out)))
  out
}
```

# Parse function 

```{r}

clean_and_parse_json <- function(text) {
  if (is.null(text) || is.na(text) || !nzchar(text)) return(tibble())
  s <- str_trim(str_remove_all(text, "```[A-Za-z]*"))
 
  # first path for when the whole response is a valid array
  if (isTRUE(validate(s))) {
    parsed <- fromJSON(s, 
                       simplifyDataFrame = TRUE)
    if (is.data.frame(parsed) && nrow(parsed)) return(as_tibble(parsed))
  }
 
  # fallback: salvage complete objects from prose or truncated output
  objs <- str_extract_all(s, r"(\{(?:[^{}"]|"(?:[^"\\]|\\.)*")*\})")[[1]]
  if (!length(objs)) return(tibble())
  map(objs, \(o) tryCatch(as_tibble(fromJSON(o)), 
                          error = \(e) NULL)) |>
    compact() |> 
    bind_rows()
}
 
# Match parsed response to original activity list 
match_to_library <- function(picks, presented) {
  if (!nrow(picks) || !"activity_id" %in% names(picks)) return(tibble())
  # fill keys the model may have omitted entirely
  for (col in c("activity", "reason")) {
    if (!col %in% names(picks)) picks[[col]] <- NA_character_
  }
 
  key <- \(x) str_squish(tolower(str_replace_all(x,
                                                 "[^A-Za-z0-9 ]", " ")))
  lookup <- set_names(presented$activity_id, 
                      key(presented$activity))
 
  picks |>
    # Keep only selected columns
    transmute(activity_id_returned = as.character(activity_id),
              activity_returned = as.character(activity),
              reason = as.character(reason),
              activity_id = coalesce(
                if_else(activity_id_returned %in% presented$activity_id,
                        activity_id_returned, 
                        NA_character_),
                unname(lookup[key(activity_returned)])),
              # T/F for if the activity ID exists
              matched = !is.na(activity_id)) |>
    left_join(presented, by = "activity_id") |>
    left_join(select(activities, activity_id, 
                     topic, actor, level),
              by = "activity_id")
}
 
```


## Response paths for runs
```{r}
out_paths <- function(stage = c("pilot", "main", "retry")){
    stage <- match.arg(stage)
    save_path <- if (stage == "main") "" else paste0("_", stage)
    list(runs = paths[[paste0("runs", save_path)]],
         raw = paths[[paste0("raw", save_path)]])
  }
```


## Extraction Loop (save raw responses)

```{r}
set.seed(m_seed)

# change replace_data to TRUE if new data is required (files will be overwritten)
execute_runs <- function(grid, stage, replace_data = FALSE) {
  stage_run <- out_paths(stage)
  # half written file name 
  check_pts <- paste0(stage_run$raw, ".unfin")
  
# Check replace_data = FALSE and if finished file exists
  if (!replace_data && file.exists(stage_run$raw)) {
    # read finished file
    saved <- readRDS(stage_run$raw)
    # extract run_ids from the finished file
# should be the complete list (pilot n = 3; main n = 1440 and failed n if any) as the following code saves to '.unfin' until grid is complete.
    saved_ids <- map_chr(saved, "run_id", 
                         .default = NA_character_)
    # check all run_id's in the grid is also in the loaded file
    if (all(grid$run_id %in% saved_ids)) {
      # if condition is met - display message
      message("All ", nrow(grid), 
              " runs already saved in ",
              stage_run$raw, 
              " (loading data, no API calls made).")
      # then load completed file and stop
      return(list(metadata = read_csv(stage_run$runs, 
                                      show_col_types = FALSE),
                  raw = saved))
    }
  }
  
# If completed file does not exist - continue w/ API calls
  # will also run if file does not contain all run_ids (should not happen with the .unfin unless grid is updated to > n) or if replace_data is TRUE. 
  results <- if (file.exists(check_pts)) readRDS(check_pts) 
                else list()
  ids_done <- map_chr(results, 
                      "run_id", 
                      .default = NA_character_)
  ids_to_do <- grid |> 
    filter(!run_id %in% ids_done)
  
  # check how many left 
  message(length(results), " completed, ", 
          nrow(ids_to_do), " remaining")
  
  for (i in seq_len(nrow(ids_to_do))) {
    row <- ids_to_do[i, ]
    
    # Prepare shuffled list and user prompt for this run
    shuffled_list <- render_list(activities_send, 
                                 seed = row$presentation_seed)
    activity_list <- format_list_for_prompt(shuffled_list)
    user_prompt <- build_user_prompt(row, activity_list)
    
    res <- tryCatch({
      
    # Build and execute request
    req <- build_request(
      provider = row$provider,
      model_id = row$model_id,
      system_prompt = row$system_prompt,
      user_prompt = user_prompt,
      params = row$params[[1]])
      
      resp <- req_perform(req)
      parsed_resp <- parse_api_response(resp, row$provider)
      list(run_id = row$run_id, 
           status = "success", 
           raw_json = resp_body_string(resp), 
           text = parsed_resp$text,
           ok = nzchar(parsed_resp$text), 
           stop_reason = parsed_resp$stop_reason %||% NA_character_,
           input_tokens = parsed_resp$in_tok %||% NA_integer_,
           output_tokens = parsed_resp$out_tok %||% NA_integer_,
           error = NA_character_)
    }, error = function(e) { # Identify why if there is an error
      list(run_id = row$run_id, 
           status = "failed",
           raw_json = NA_character_, 
           text = NA_character_,
           ok = FALSE,
           stop_reason = NA_character_, 
           input_tokens = NA_integer_,
           output_tokens = NA_integer_, 
           error = conditionMessage(e))
    })
    # save res to next line in results (check_pts - .unfin)
    results[[length(results) + 1L]] <- res
    # save every call so no data is lost 
    saveRDS(results, check_pts)
    
    # track progress and if there are any issues
    message(row$run_id, " | ", 
            row$model, " : ", 
            res$status, " | ",
            coalesce(res$stop_reason %||% NA_character_,
                     ""))
    
    Sys.sleep(req_pause)
  }
  # Save metadata 
  metadata <- grid |> 
    select(-params, -template, -system_prompt) |> 
    left_join(select(bind_rows(results), 
                     -raw_json, -text), 
              by = "run_id")
  
  write_csv(metadata, stage_run$runs)
  saveRDS(results, stage_run$raw)
  if(file.exists(check_pts)) file.remove(check_pts)
  list(metadata = metadata, 
       raw = results)
}

```

# Run 1. Pilot Run 
```{r}
#| eval: false

set.seed(m_seed)
# smaller trial than S1 - just to the three models
pilot_grid <- grid_run |> 
  slice_sample(n = 1,
               by = c(model))

pilot_run <- execute_runs(pilot_grid, 
                          stage = "pilot")
```



### Check Results 

```{r}
# 
pilot_raw  <- readRDS(paths$raw_pilot)
pilot_meta <- read_csv(paths$runs_pilot, 
                       show_col_types = FALSE)

# Check that all runs are present 
nrow(pilot_meta) # should = 3
# Check that runs were successful 
pilot_meta |> count(status, ok) # success/TRUE
# Check that runs were successful (STOP, completed or end_turn) or if they stopped (e.g., max_tokens or  incomplete) 
pilot_meta |> count(stop_reason) 


# check clean_and_parse_json function 
main_check <- bind_rows(pilot_raw) |>
  select(run_id, status, ok, 
         stop_reason, input_tokens, 
         output_tokens, text) |>
  left_join(select(pilot_meta, run_id, 
                   model, instruction_type, 
                   condition, k),
            by = "run_id") |>
  mutate(
    # check that response begins with [ (no preamble)
    starts_json = str_starts(str_trim(text), fixed("[")),
    # check that response ends with ] (no commentary or truncated mid response)
    ends_json = str_ends(str_trim(text), fixed("]")),
    # check if response is a valid JSON 
    valid_clean = map_lgl(text, \(t) isTRUE(validate(str_trim(
      # remove markdown fences 
      str_remove_all(t, "```[A-Za-z]*"))))),
    # check k using "activity_id" 
    n_id_keys = str_count(text, fixed('"activity_id"')),
    # check how many rows the parse function extracted
    n_parsed = map_int(text, \(t) nrow(clean_and_parse_json(t))))
 
# Check if the clean_and_parse_json function had to extract JSON responses per-object due to raw parse fail.
pilot_check |>
  count(model, starts_json, 
        ends_json, valid_clean,
        recovered = n_parsed > 0)
 
# Any response where objects were lost between the text and the parse
pilot_check |>
  filter(n_parsed != n_id_keys) |>
  select(run_id, model, condition, 
         n_id_keys, n_parsed, stop_reason)
 
# Check if token n set is okay
# Does not include all tokens for gemini - but gemini maxes at 36000 and is currently set to 32000 
pilot_check |>
  summarise(min_out = min(output_tokens), 
            med_out = median(output_tokens),
            max_out = max(output_tokens), 
            .by = c(model, condition)) |>
  mutate(tokens = max_tokens - max_out) |>
  arrange(tokens)

```

#Run 2. Main Run 
```{r}
#| eval: false
set.seed(m_seed)
# Run the execution
main_run <- execute_runs(grid_run, stage = "main")
```

### Selections 

Save selections_long once after `main_run` or continue to **Load**
Create and save data frame with one row for each activity returned (~ 15 rows per run_id).

```{r}
#| eval: false
selections_long <- bind_rows(main_run$raw) |>
  # select runs that were successful w/ response
  filter(status == "success", ok) |>
  select(run_id, text) |>
  left_join(main_run$metadata, by = "run_id") |>
  mutate(picks = map2(text, presentation_seed,
                      \(t, s) match_to_activities(
                        clean_and_parse_json(t),
                        render_list(activities_send, s)))) |>
  select(-text) |>
  unnest(picks, 
         keep_empty = TRUE)

write_csv(selections_long, 
          paths$parsed)
```


## Load 

Load data if the above code has not been run

```{r}
# Load data 
main_raw <- readRDS(paths$raw)
main_meta <- read_csv(paths$runs,
                       show_col_types = FALSE)
selections_long <- read_csv(paths$parsed,
                       show_col_types = FALSE)
```

##Check Results

### Functions 

```{r}

# 1 - Check that all runs are present 
nrow(main_meta) # should = 1440
# 2 - Check that runs were successful 
main_meta |> count(status, ok) # success/TRUE
# 3 - Check that runs were successful (STOP, completed or end_turn) or if they stopped (e.g., max_tokens or  incomplete) 
main_meta |> count(stop_reason)  
# 4 - check that all run_ids are present
n_distinct(selections_long$run_id) # 1440

# 5 - check clean_and_parse_json function 
main_check <- bind_rows(main_raw) |>
  select(run_id, status, ok, 
         stop_reason, input_tokens, 
         output_tokens, text) |>
  left_join(select(main_meta, run_id, 
                   model, instruction_type, 
                   condition, k),
            by = "run_id") |>
  mutate(
    # check that response begins with [ (no preamble)
    starts_json = str_starts(str_trim(text), fixed("[")),
    # check that response ends with ] (no commentary or truncated mid response)
    ends_json = str_ends(str_trim(text), fixed("]")),
    # check if response is a valid JSON 
    valid_clean = map_lgl(text, \(t) isTRUE(validate(str_trim(
      # remove markdown fences 
      str_remove_all(t, "```[A-Za-z]*"))))),
    # check k using "activity_id" 
    n_id_keys = str_count(text, fixed('"activity_id"')),
    # check how many rows the parse function extracted
    n_parsed = map_int(text, \(t) nrow(clean_and_parse_json(t))))

# 6 - Check if the clean_and_parse_json function had to extract JSON responses per-object due to invalid JSON document.
main_check |>
  count(model, starts_json,
        ends_json, valid_clean,
        recovered = n_parsed > 0)

# 7 - Check if number of sentences extracted by the parse function does not match k ()
main_check |>
  filter(n_parsed != n_id_keys) |>
  select(run_id, model, 
         condition, n_id_keys, 
         n_parsed, stop_reason)

```

### Data Summary  

```{r}
# 1 - Main summary table 
main_summary <- selections_long |>
  summarise(
    # check no NAs in returned IDs
    n_returned = sum(!is.na(activity_id_returned)),
    # check all IDs were matched in the selections_long df
    n_matched = sum(matched, na.rm = TRUE),
    n_unmatched = sum(!matched, na.rm = TRUE),
    # check if model sent back duplicate IDs in one run
    n_duplicated = sum(duplicated(
      na.omit(activity_id))),
    .by = c(run_id, model, 
            instruction_type, 
            condition, k)) |>
  # check if all fixed_k runs returned 15 
  mutate(k_ok = is.na(k) | n_returned == k,
  # check if a run returned < k = 1
         nonzero = n_returned > 0) |>
  arrange(model, condition)

# 2 - Check all runs returned 15 
main_summary |> filter(!k_ok)

# 3 - Check if any k = 0
main_summary |> filter(!nonzero)

# 4 - Check for number of unmatched IDs 
main_summary |> filter(n_unmatched >= 1)

# 4.1 - only run if unmatched n > 0 
#     - Check the IDs returned not in the library
# selections_long |>
#   filter(!matched) |>
#   select(run_id, model, 
#          activity_id_returned, 
#          activity_returned)

# 5 - check for number of duplicated selections
main_summary |> filter(n_duplicated > 0)

# 5.1 - only run if n_duplicated n > 0
# duplicated_sel <- main_summary |>
#   filter(n_duplicated > 0) |>
#   pull(run_id)
# # Duplicated rows - original and returned IDs and activities
# selections_long |>
#   filter(run_id %in% dupe_runs) |>
#   filter(activity_id %in% activity_id[duplicated(activity_id)],
#          .by = run_id) |>
#   select(run_id, activity_id_returned,
#          activity_id, activity_returned,
#          activity, reason, model)

```

### Response vs Original pairs

```{r}
# 1 - check that IDs and activity sentences match check activity sentences and IDs from model match original list
mismatches <- selections_long |>
  # agrees = TRUE when the activity is the same as the one the ID matched to
  mutate(agrees = norm(
    activity_returned) == norm(activity)) |>
  count(model, agrees)

mismatches

# 2 - If FALSE - check the activities that do not match original list 
selections_long |>
  # filter returned activities not matched to original activity 
  filter(norm(activity_returned) != norm(activity)) |>
  select(run_id, model, 
         activity_id, 
         activity_returned, 
         activity) |> 
 # count the number of times exact same mismatches occur 
         count(activity_id, 
                        activity_returned, 
                        activity,
               name = "n_runs", sort = TRUE) |> 
  # print mismatched pairs and n for readability
  with(paste(activity_id, "(", n_runs, "runs)\n",
            "model:", activity_returned, "\n",
            "orig.:", activity, "\n\n")) |> 
  cat()

# 3 - Check whether any of the mismatched pairs match a different activity ID 
selections_long |>
  # create column that prints the activity_ID for the original activity
  mutate(text_match_id = activity_id[
    # match finds the activity that matches to the activity returned
    match(norm(activity_returned), 
          norm(activity))]) |>
  select(run_id, activity_id, 
         text_match_id, model,
         reason) |> 
  # remove activity ids that match 
  filter(activity_id != text_match_id)


# RESULT - One run matched a different activity - run_id and activity recorded to hardcode in the `Stage_1.qmd` data prep. 
```


# Run 3. Check for failed runs 
Should not be any present without it having been flagged above, but just to check.

```{r}
retry_ids <- main_run$metadata |>
  filter(!coalesce(ok, FALSE) | 
          coalesce(status, "missing") != "success" |
          coalesce(stop_reason, "") %in% c("max_tokens", 
                                           "MAX_TOKENS", 
                              "max_output_tokens", 
                              "incomplete")) |>
  pull(run_id)

retry_ids
# NONE 
```

### Retry failed runs if any

Would need to join to main data and re-save. 

```{r}
#| eval: false 

set.seed(m_seed)

retry_grid <- grid_run |> 
  filter(run_id %in% retry_ids)

retry_run <- execute_runs(retry_grid, stage = "retry")
```





