---
title: "Stage_2"
format: html
---

# Load Packages 

```{r}
library(geepack)
library(emmeans) 
library(dplyr)
library(tidyr)
library(sandwich)
library(readr)
library(stringr)
```


# Load data

```{r}
activities <- read_csv("data/activity_main.csv", 
                       show_col_types = FALSE)

selections_long <- read_csv("response/selections_long_s2.csv",
                            show_col_types = FALSE)
```

# Data handling

Checks in `API_data_s2.qmd` identified:

**Four** GTP-Terra runs returned one activity twice, each time with an identical reason indicating repetition rather than deliberate selection.
* Duplicates were collapsed and only counted once (K_u = 14 for run_080, run_218, run_399 and run_407).

**One** sonnet5 run returned 16 activity selections. 
* All 16 selections for the run were retained, as selection order carries no information under the output instructions (K_u = 16 for run_564).

76 selections returned activity sentences that differed from the activity list provided. 72 of which were minor rewrites of the same activity ID. 
* These minor mismatched sentences were determined to be okay to left as is. 

**Four** of the 76 mismatched sentences from Sonnet 5 returned activity text for a different activity ID. No mismatched sentences were in the same axis category and the reasons provided for each selection were too vague to determine the intended selection. 
* All four mismatched sentences were excluded due to their ambiguity (K_u = 14 for run_492, run_589, run_732 and run_753) 

**One** sonnet5 response was an invalid JSON but was fully recovered by per-object parsing (k = 15 for RUN?????????????????????????????).
And why it was invalid Code: bad <- main_check |> filter(!valid_clean)
attr(validate(str_trim(str_remove_all(bad$text, "```[A-Za-z]*"))), "err") 

The Poisson offset `log(K_u)` accommodates the five runs where K_u ≠ 15.

```{r}
# normalise and clean text string in case of minor changes 
# removes random spaces, makes everything lowercase
norm <- function(x) {
  x |> 
    # remove puntuation, new lines etc. 
    str_replace_all("[^A-Za-z0-9 ]", " ") |> 
    str_to_lower() |> 
    str_squish()
}

selections_long <- selections_long |>
  # collapse duplicates w/ identical reasoning
  distinct(run_id, activity_id,
           .keep_all = TRUE) |> 
  mutate(text_match_id = activity_id[
    # match finds the activity that matches to the activity returned
    match(norm(activity_returned), 
          norm(activity))],
    # keep the 72 NAs, remove only the four ambiguous one
    remove = !is.na(text_match_id) & activity_id != text_match_id) |> 
    filter(!remove) |> 
  select(c(-remove, -text_match_id))

# Check K after cleaning
selections_long |>
  count(run_id, name = "K") |> # selections per run after cleaning
  count(K) # how many runs have each K

```

# Actor

##Build data frame

One row per run x category, with zero counts filled in. Reference levels are the largest categories from Stage 1 (upstream, low_cog, and upstream_low_cog, the largest cell in all nine strata), which gives the smallest SEs on every rate ratio. Rows are sorted by `run_id`, which `geeglm` requires.

```{r}

# one row per selected activity, with actor mapped
data_actor <- selections_long |> 
  # count produces rows for comibinations 
  count(run_id, model, 
        instruction_type, actor, 
        name = "y") |>
  # complete - adds rows missing (i.e., not selected in a run)
  # nesting - add missing row for every run_id, model and instruction_type combination 
  complete(nesting(run_id, model, 
                   instruction_type), 
           actor, 
           # set missing to 0 (rather than NA)
           fill = list(y = 0)) |>
  # geeglm requires clusters to be sequenced
  group_by(run_id) |> 
  mutate(K = sum(y)) |> 
  ungroup() |>
  mutate(actor = relevel(factor(actor),
                         ref = "upstream"),
         model = factor(model), 
         instruction_type = factor(instruction_type),
         # geeglm requires ordered clusters as integers
         run_id = as.integer(factor(run_id))) |>
  arrange(run_id, actor)

```


# 4.2 Selection Patterns
analyse run–actor counts with a run–clustered Poisson GEE using actor effects and actor×(model, instruction type) interactions (no intercept). We do not include model×instruction type (or higher-order) interactions, so prompt contrasts are averaged over models and vice versa. We report:

(i)	actor rate ratios (RRs) relative to Event, which we use as an arbitrary reference category, and; 
(ii)	within–actor pairwise RRs for model and for instruction type.

##1. Fit Poisson GEEs

Poisson GEEs with `offset(log(K_u))`, clustered by run, with robust (sandwich) SEs. The primary specification uses an independence working correlation. 


```{r}
# Poisson GEE - no intercept - exposure offset log K 
# actor, actor*model, actor*intruction effects
# prompt contrasts averages over models 
# Sandwich SEs are default (std.err = "san.se")
fit_actor <- geeglm(y ~ 0 + actor + actor:model + actor:instruction_type + offset(log(K)),
                 family = poisson, 
                 data = data_actor, 
                 #cluster by run 
                 id = run_id,
                 # correlation structure: exchangeable
                 # all obs within the cluster (run_id) share equal correlation 
                 corstr = "independence")

```

##2. Model Adequacy 

```{r}
# Pearson chi2 / residual degrees of freedom 
sum(residuals(fit_actor, 
              type = "pearson")^2) / df.residual(fit_actor)  
# deviance/df
glm_actor <- glm(formula(fit_actor), 
                 family = poisson, 
                 data = data_actor)

# gives a dispersion parameter estimate to check if the independence working correlation assumption is valid
deviance(glm_actor) / df.residual(glm_actor)

```

##3. Build Reference Grid 
averages over model and instruction type. offset = 0 returns rates per unit of exposure. The ratios are the same either way, because the offset cancels in contrasts.


```{r}
emm_actor <- emmeans(fit_actor, ~ actor, 
                     offset = 0, 
                     data = data_actor)
```


##4. Wald contrasts vs. Upstream (rate ratios)
Inference uses Wald χ2(1) tests on log–rate contrasts with robust covariance, and 95% CIs are exp(ˆθ ± 1.96 SE). Tables indicate the number of run clusters. We report pairwise differences only when BH–FDR q < .05 and |∆%| ≥ 10.

(i)	actor rate ratios (RRs) relative to upstream, which we use as an arbitrary reference category, and 
(ii)	within–actor pairwise RRs for model and for instruction type.


```{r}
table_4 <- contrast(emm_actor, 
                    method = "trt.vs.ctrl", 
                    ref = 1, 
                    adjust = "none") |>
  summary(infer = TRUE, type = "response")
# columns: ratio (= RR), asymp.LCL, asymp.UCL, z.ratio (Wald chi2 = z.ratio^2), p.value
table_4

# w/ midstream 
pairs(emm_actor, 
      adjust = "none") |>
  summary(infer = TRUE, 
          type = "response")
```

### Cross-Model stability in actor Preferences 
```{r}
model_act <- emmeans(fit_actor, ~ model | actor, 
                   offset = 0, 
                   data = data_actor) |>
  pairs(adjust = "none") |>
  summary(infer = TRUE, 
          type = "response") |>
  as_tibble() |>
  mutate(q = p.adjust(p.value, 
                      method = "BH"), # Step 5: BH
         delta = (ratio - 1) * 100) # Delta% = (RR - 1) x 100

model_act

table_5 <- emmeans(fit_actor, ~ actor | model, 
                   offset = 0,
                   nesting = NULL) |>
  summary(type = "response")
  
table_5

```

### Instruction stability in actor preference
```{r}
ins_diff <- emmeans(fit_actor, ~ instruction_type | actor, 
                   offset = 0, 
                   data = data_actor) |>
  pairs(adjust = "none") |>
  summary(infer = TRUE, 
          type = "response") |>
  as_tibble() |>
  mutate(q = p.adjust(p.value, 
                      method = "BH"), # Step 5: BH
         delta = (ratio - 1) * 100) # Delta% = (RR - 1) x 100

ins_diff

```


# Levels H,L Analysis 
Category-level analyses use four fits, one per element, on run × category rows with offset log(K_elem), dropping runs where K_elem = 0.

## Build Data frame 


```{r}

df_levels <- selections_long |> 
    mutate(actor = factor(actor, 
                          levels = c("upstream", 
                                     "midstream", 
                                     "downstream")),
           level = factor(level, 
                          levels = c("high_cog", 
                                     "low_cog"))) |>
  # count produces rows for comibinations 
  count(run_id, model, instruction_type, 
        actor, level, 
        name = "y") |>
  # complete - adds rows missing (i.e., not selected in a run)
  # nesting - add missing row for every run_id, model, instruction_type across every actor and level combination 
  complete(nesting(run_id, model, instruction_type), 
           # cross actor*level
           actor, level, 
           # set missing to 0 (rather than NA)
           fill = list(y = 0)) |>
  # run budget for actor-level (4.2) model
  group_by(run_id) |> 
  mutate(K = sum(y)) |>
  # run budget for within-actor (4.3) model
  group_by(run_id, actor) |>
  mutate(K_actor = sum(y)) |>            
  ungroup() |>
  mutate(model = factor(model), 
         instruction_type = factor(instruction_type),
         run_id = as.integer(factor(run_id)))


data_levels <- df_levels |>
  # offset log(0) = -Inf; rows carry no information
  filter(K_actor > 0) |>
  # geeglm needs each run's rows contiguous
  arrange(run_id, actor, level)

```


# Actor*level actual 
To compare cells across actors, use a 6-cell Poisson GEE with offset log(K), the run's total budget. 

```{r}
data_cells <- df_levels |> 
  arrange(run_id, actor, level)
stopifnot(all(data_cells$K > 0))   # drop any zero-selection runs first; log(0) = -Inf

# Fit GEE
fit_cell <- geeglm(y ~ 0 + actor:level + actor:level:model + actor:level:instruction_type + offset(log(K)),
                   family = poisson, 
                   data = data_cells, 
                   id = run_id, 
                   corstr = "independence")

# Model Adequacy 
# Pearson chi2 / residual degrees of freedom 
sum(residuals(fit_cell, 
              type = "pearson")^2) / df.residual(fit_cell)  
# deviance/df
glm_cells <- glm(formula(fit_cell), 
                 family = poisson, 
                 data = data_cells)

# gives a dispersion parameter estimate to check if the independence working correlation assumption is valid
deviance(glm_cells) / df.residual(glm_cells)                        




# global 
emm_cell <- emmeans(fit_cell, ~ actor * level, 
                    offset = 0, 
                    data = data_cells, 
                    nesting = NULL)

# All 15 cell-vs-cell rate ratios, averaged over model and instruction
tab_cells <- pairs(emm_cell, 
                   adjust = "none") |>
  summary(infer = TRUE, 
          type = "response") |>
  as_tibble() |>
  mutate(q = p.adjust(p.value, method = "BH"),
         delta = (ratio - 1) * 100)
# ratio follows the label order, e.g. "upstream low_cog / downstream high_cog"; 1/ratio flips it
tab_cells



```


```{r}
emm_act_lev <- emmeans(fit_cell, ~ level * actor, 
                       offset = 0,
                       data = data_cells, 
                       nesting = NULL)

# 1. Within-actor split (low_cog / high_cog in each actor)
tab_split <- contrast(emm_act_lev, 
                      method = "trt.vs.ctrl", 
                      by = "actor", 
                      adjust = "none") |>
  summary(infer = TRUE, 
          type = "response")


tab_split

# check if cog complexity differ across actors
ic_actor <- contrast(emm_act_lev, 
                     interaction = c("trt.vs.ctrl", "pairwise"), 
                     adjust = "none")
test(ic_actor, joint = TRUE) # omnibus Wald test, df1 = 2

tab_actor_int <- summary(ic_actor, 
                         infer = TRUE, 
                         type = "response") |>
  as_tibble() |>
  mutate(q = p.adjust(p.value, 
                      method = "BH"))
tab_actor_int


# Check if the cog complexity selection differs across models / instructions within each actor?
split_by <- function(fac) {
  ic <- emmeans(fit_cell, 
                as.formula(paste("~ level *", fac, "| actor")),
                offset = 0, 
                data = data_cells, 
                nesting = NULL) |>
    contrast(interaction = c("trt.vs.ctrl", "pairwise"), 
             by = "actor", 
             adjust = "none")
  list(omnibus = test(ic, joint = TRUE, 
                      by = "actor"),
       pairs = summary(ic, infer = TRUE, 
                       type = "response") |>
         as_tibble() |> 
         mutate(q = p.adjust(p.value, 
                             method = "BH")))
}
fig_split_model <- split_by("model")# 3 actors x 3 model pairs
split_instr <- split_by("instruction_type")  # 3 actors x 3 instruction pairs
fig_split_model
split_instr

```

## Graph 
```{r}

library(ggplot2)
library(scales)
 
thr <- 1.50   # RR threshold used for the split contrasts
 
model_labs <- function(x) {
  x <- str_to_lower(str_remove_all(x, "[()]"))
  case_when(str_detect(x, "gemini") ~ "Gemini Flash-3.8",
            str_detect(x, "gpt") ~ "GPT-Terra",
            str_detect(x, "sonnet") ~ "Sonnet 5",
            TRUE ~ x)
}
 
relabel_pair <- function(lab) {
  parts <- str_split_fixed(lab, " / ", 2)
  paste(model_labs(parts[, 1]), "/", model_labs(parts[, 2]))
}
 
actor_labs <- c(upstream = "Upstream (government)",
                midstream = "Midstream (community)",
                downstream = "Downstream (individual)")
 
fig_df <- tab_split_model$pairs |>
  rename(any_of(c(lower.CL = "asymp.LCL",
                  upper.CL = "asymp.UCL"))) |>
  rename(pair = starts_with("model")) |>
  mutate(pair = relabel_pair(as.character(pair)),
         actor = factor(actor_labs[as.character(actor)], 
                        levels = actor_labs),
         status = case_when(q < .05 & (ratio >= thr | ratio <= 1/thr) ~ "Significant, beyond threshold",
                            q < .05                                    ~ "Significant, within threshold",
                            TRUE                                       ~ "Not significant"),
         status = factor(status, 
                         levels = c("Significant, beyond threshold",
                                    "Significant, within threshold",
                                    "Not significant")),
         pair = factor(pair, 
                       levels = rev(unique(pair))))
 
stopifnot(!anyNA(fig_df$actor))
 
ggplot(fig_df, aes(x = ratio, 
                   y = pair, 
                   colour = status)) +
  annotate("rect", 
           xmin = 1/thr,  xmax = thr, 
           ymin = -Inf, ymax = Inf,
           fill = "grey92") +
  geom_vline(xintercept = 1, 
             linetype = "dashed",
             colour = "grey55", 
             linewidth = .4) +
  geom_linerange(aes(xmin = lower.CL, 
                     xmax = upper.CL), 
                 linewidth = 1.1) +
  geom_point(size = 2.8) +
  geom_text(aes(x = upper.CL, 
                label = number(ratio, 
                               accuracy = .01)),
            hjust = -.25, 
            size = 2.8, 
            show.legend = FALSE) +
  # actor names above each panel
  facet_wrap(~ actor, ncol = 1, scales = "free_y") +
  scale_x_log10(breaks = c(.25, .5, 1, 2, 4),
                labels = c("0.25", "0.5", "1", "2", "4"),
                expand = expansion(mult = c(.05, .12))) +
  # drop = TRUE removes empty legend keys (e.g. no non-significant contrasts)
  scale_colour_manual(values = c("Significant, beyond threshold" = "#D55E00",
                                 "Significant, within threshold" = "#0072B2",
                                 "Not significant" = "grey60"),
                      drop = TRUE) +
  labs(x = "Ratio of low / high splits between models (log scale, 95% CI)",
       y = NULL, colour = NULL,
       title = "Cognitive Complexity Differs Between Models",
       subtitle = paste0("Values above 1: first model's split is more low-weighted.\n",
                         "Shaded band: RR 1/", thr, " to ", thr, " threshold."),
       caption = "q values BH-adjusted across the nine contrasts. Estimates average over instruction type.") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top",
        legend.justification = "left",
        # align title, subtitle, legend and caption with the whole plot, not the panel
        plot.title.position = "plot",
        plot.caption.position = "plot",
        strip.text = element_text(face = "bold", 
                                  hjust = 0, 
                                  size = 10),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        panel.spacing.y = unit(.8, "lines"),
        plot.title = element_text(face = "bold", 
                                  size = 13),
        plot.subtitle = element_text(colour = "grey35", 
                                     size = 9),
        plot.caption = element_text(colour = "grey45", 
                                    size = 8, hjust = 0),
        plot.margin = margin(8, 16, 8, 8))
 
ggsave("fig_table8_split_contrasts.png", width = 7.5, height = 6.5, dpi = 300)

```



