---
title: "Stage_1_actual"
format: html
---


# Load Packages

```{r}
library(dplyr)
library(tidyr)
library(readr)
library(sandwich)
library(lmtest)
library(emmeans)
```


# Load data

```{r}
activities <- read_csv("data/activity_main.csv", 
                       show_col_types = FALSE)

selections_long <- read_csv("response/selections_long.csv",
                            show_col_types = FALSE)
```

# Data handling

Checks in `API_data_s1.qmd` identified:

29 selections returned activity sentences that differed from the activity list provided, 28 of which were minor rewrites of the same activity ID. 
* These minor mismatched sentences were retained as is.

**One** of the 29 mismatched sentences (sonnet5) returned activity text for a different activity ID. No mismatched sentence was in the same axis cell and the reason provided was too vague to determine the intended selection. 
* The selection was excluded due to its ambiguity (K_u = 14 for run_356, fixed budget).

All 270 fixed budget runs returned exactly 15 selections before this exclusion. Selection shares (and the `log(K_u)` offset in the Poisson fits) accommodate run_356.

```{r}
selections_long <- selections_long |>
  # remove the one ambiguous selection
  filter(!(run_id == "run_356" & activity_id == "a_DN06"))

# fixed budget compliance after cleaning
selections_long |>
  filter(condition == "k_fixed") |>
  # selections per run after cleaning
  count(run_id, name = "K") |> 
  count(K) 
```

# Build data frame

One row per run x cell, with zero counts filled in. The six cells are the actor x level combinations. `K_u` is the number of selections in a run and `s_uc = y_uc / K_u` is the within-run selection share, so the six shares sum to one in every run. `k_free` is the reference level of `condition`, so every contrast reads fixed minus free.

```{r}
# the six outcome cells (actor x level)
axes <- activities |>
  distinct(actor, level) |>
  mutate(axis_cell = paste(actor, level, 
                           sep = "_")) |>
  arrange(axis_cell)

sel_long <- selections_long |>
  mutate(axis_cell = paste(actor, level, 
                           sep = "_")) |>
  # count produces rows for combinations 
  count(run_id, condition, model, 
        instruction_type, axis_cell, 
        name = "y_uc") |>
  # complete - adds rows missing (i.e., cell not selected in a run)
  # nesting - add missing row for every run_id, condition, model and instruction_type combination
  complete(nesting(run_id, condition, model, 
                   instruction_type), 
           axis_cell = axes$axis_cell, 
           # set missing to 0 (rather than NA)
           fill = list(y_uc = 0)) |>
  group_by(run_id) |>
  # K_u = selections in the run, s_uc = share of the run's selections in each cell
  mutate(K_u = sum(y_uc), 
         s_uc = y_uc / K_u) |>
  ungroup() |>
  mutate(condition = factor(condition, 
                            levels = c("k_free", "k_fixed")),
         axis_cell = factor(axis_cell),
         model = factor(model),
         instruction_type = factor(instruction_type))

```


# 4.1 Comparison of Experimental Setups

Condition contrasts compare cell shares under the fixed (K = 15) and free budget. The free budget K is described first because it decides how the two arms can be weighted.

## 1. Selections under a free budget (Table 1)

If most of the variance in free budget K lies between models, K-weighting reweights models rather than buying precision (K-weights hand sonnet5 about half the free arm against a third of the fixed arm). Unweighted estimates are therefore primary. Each model contributes 90 runs per arm, so unweighted = model-balanced.

```{r}
# one row per free budget run
runs_free <- sel_long |>
  distinct(run_id, condition, model, K_u) |>
  filter(condition == "k_free")

table_1 <- runs_free |>
  group_by(model) |>
  summarise(mean_K = mean(K_u),
            sd_K = sd(K_u),
            # modal K (ties joined) and % of runs at the mode
            mode_K = paste(names(which(table(K_u) == max(table(K_u)))), 
                           collapse = " and "),
            mode_pct = 100 * max(table(K_u)) / n(),
            min_K = min(K_u),
            max_K = max(K_u),
            total_K = sum(K_u)) |>
  # share of all free arm selections (fixed arm = 33.3% each)
  mutate(free_weight = 100 * total_K / sum(total_K),
         # change imposed by the fixed budget of 15
         forced_change = 15 - mean_K)
table_1

# between-model share of the variance in free budget K
aov_K <- aov(K_u ~ model, 
             data = runs_free)
summary(aov_K)

# eta^2 = between-model sum_sq / total sum_sq
sum_sq <- summary(aov_K)[[1]][["Sum Sq"]]
sum_sq[1] / sum(sum_sq)
```

## 2. Fit condition model

Condition contrasts use least squares on selection shares with run-clustered SEs. Each run contributes six rows whose shares sum to one, so the rows are correlated within a run. `condition * axis_cell` gives a separate condition effect for every cell, and heteroskedasticity is handled by the sandwich (no weighting needed for valid inference).

```{r}
# OLS on shares
fit_cond <- lm(s_uc ~ condition * axis_cell, 
               data = sel_long)

# run-clustered sandwich 
# HC0 without small-sample adjustment, same as the GEE sandwich in Stage 2
V_cond <- vcovCL(fit_cond, 
                 cluster = ~ run_id, 
                 type = "HC0", 
                 cadjust = FALSE)

```

## 3. Build Reference Grid

`df = Inf` gives Wald z-tests and 95% CIs as estimate ± 1.96 SE.

```{r}
emm_cond <- emmeans(fit_cond, ~ condition | axis_cell, 
                    vcov. = V_cond, 
                    df = Inf)
```

## 4. Wald contrasts, fixed vs. free (Table 2)

Estimates are percentage points (fixed minus free). BH-FDR is applied across the six cells as one family. Because shares sum to one within a run, the six contrasts sum to zero by construction. The K-weighted contrast (`K_weight`) is reported alongside for comparison.

```{r}
# free and fixed shares per cell (%), unweighted (primary) and K-weighted
shares <- sel_long |>
  summarise(pct = 100 * mean(s_uc),
            # K-weighted = pooled share of all selections
            pct_K = 100 * sum(y_uc) / sum(K_u),
            .by = c(condition, axis_cell)) |>
  pivot_wider(names_from = condition, 
              values_from = c(pct, pct_K)) |>
  mutate(K_weight = pct_K_k_fixed - pct_K_k_free)

table_2 <- contrast(emm_cond, 
                    method = "trt.vs.ctrl", 
                    ref = 1, 
                    adjust = "none") |>
  summary(infer = TRUE) |>
  as_tibble() |>
  # proportions to percentage points
  mutate(across(c(estimate, SE, asymp.LCL, asymp.UCL), 
                \(x) 100 * x),
         q = p.adjust(p.value, 
                      method = "BH")) |> # BH across the six cells
  left_join(shares, 
            by = "axis_cell")
# columns: estimate (pp), asymp.LCL, asymp.UCL, z.ratio (Wald chi2 = z.ratio^2), p.value, q, unweighted (pct) and K-weighted (pct_K) shares, delta_K
table_2 |> select(axis_cell, 
                  estimate,
                  q, asymp.LCL, 
                  asymp.UCL,
                  pct_k_free,
                  pct_k_fixed)
# For Annex 
table_2 |> select(K_weight,
                  pct_K_k_free,
                  pct_K_k_fixed)
```

### Heterogeneity across models and instruction types (Table 3)

```{r}
fit_het <- lm(s_uc ~ condition * axis_cell * model + condition * axis_cell * instruction_type,
              data = sel_long)

# run clustered sandwich SEs 
V_het <- vcovCL(fit_het, 
                cluster = ~ run_id, 
                type = "HC0", 
                cadjust = FALSE)

# omnibus: condition x cell x model
waldtest(update(fit_het, . ~ . - condition:axis_cell:model), 
         fit_het, 
         test = "Chisq", 
         vcov = V_het)

# omnibus: condition x cell x instruction type
waldtest(update(fit_het, . ~ . - condition:axis_cell:instruction_type), 
         fit_het, 
         test = "Chisq", 
         vcov = V_het)
```

Per axis cell to see if the condition effect differs across the three models (or instruction types).

```{r}
het_by <- function(fac) {
  emmeans(fit_het, 
          as.formula(paste("~ condition *", fac, "| axis_cell")),
          vcov. = V_het, 
          df = Inf) |>
    # (fixed - free) compared pairwise across the levels of fac
    contrast(interaction = c("trt.vs.ctrl", "pairwise"), 
             by = "axis_cell", 
             adjust = "none") |>
    test(joint = TRUE, 
         by = "axis_cell") |>
    as_tibble() |>
    mutate(chisq = qchisq(p.value, df1, 
                          lower.tail = FALSE),
           q = p.adjust(p.value, 
                        method = "BH")) |> # BH within column
    select(axis_cell, df1, chisq, p.value, q)
}

table_3 <- left_join(het_by("model"), 
                     het_by("instruction_type"),
                     by = "axis_cell", 
                     suffix = c("_model", "_instr"))
table_3 |> select(axis_cell, 
                  chisq_model,
                  q_model,
                  chisq_instr,
                  q_instr) |> 
  rename("Axis Cell" = axis_cell, 
         " χ^2 (model)" = chisq_model,
         "q (model)" = q_model,
         "χ^2  (instruction)" =chisq_instr,
         "q (instruction)" = q_instr)


```

### Model-specific condition contrasts (Table 4)

All six cells x three models, BH-FDR across the 18 contrasts as one family (fixed before looking at Table 3, so the family does not depend on which cells were heterogeneous).

```{r}
table_4 <- emmeans(fit_het, ~ condition | axis_cell * model, 
                   vcov. = V_het, 
                   df = Inf) |>
  contrast(method = "trt.vs.ctrl", 
           ref = 1, 
           adjust = "none") |>
  summary(infer = TRUE) |>
  as_tibble() |>
  mutate(across(c(estimate, SE, asymp.LCL, asymp.UCL), 
                \(x) 100 * x),
         q = p.adjust(p.value, 
                      method = "BH")) # BH across 18 contrasts
table_4 |> filter(axis_cell == c("downstream_high_cog", 
                                 "downstream_low_cog", 
                                 "upstream_low_cog")) |> 
  # filter heterogenous cells
  select(axis_cell,
         model, 
         estimate, 
         asymp.LCL,
         asymp.UCL, 
         q)
```


# Power analysis for Stage 2 replication count

For each model x instruction stratum, find the smallest R giving ≥ 80% power to detect the target RR, using that stratum's own Stage 1 composition. The Stage 2 count is the 80th percentile across the nine strata, taking the larger of the two tiers (as in Jung et al., Appendix E.1).

The simulation targets the pairwise model and instruction contrasts, where the null RR = 1 is substantive. Composition contrasts (cell vs. cell) were already resolved at Stage 1 and need no power target. Model is a moderator of the condition effect (Table 3), so R is computed per stratum rather than pooled.

## 1. Settings

* RR = 1.50 (cell tier) is roughly half the narrowest cross-cell spread in any stratum (checked in step 2), and matches the source design.
* RR = 1.20 (actor tier) aggregates six cells into three actors, so larger counts per group make a smaller RR comparably demanding.
* The level margin (two groups of three cells) is better powered than the actor margin at any RR, so it needs no tier of its own.

```{r}
# fixed budget (Stage 2 condition)
k_budget <- 15 
alpha <- 0.05
# within-stratum power target
power_t <- 0.80 
# across-strata coverage quantile
cover_q <- 0.80 
# simulations per (stratum, R)
n_sim <- 1000 
# candidate replication counts, ascending
r_grid <- seq(20, 400, by = 20) 

rr_cell <- 1.50  
rr_actor <- 1.20

cells <- axes$axis_cell
# actor of each cell, same order as cells
actor_of <- axes$actor
```

## 2. Strata inputs

One row per stratum from the fixed arm with:
* dispersion from a Poisson fit (Pearson χ²/df and deviance/df). φ ≤ 1 (no overdispersion) means the multinomial draws below are not too optimistic, so no overdispersion correction is needed.
* `spread` = largest / smallest cell share, which rr_cell should sit well below. 
* the target group = the smallest cell and the smallest actor. The hardest pairing to detect, so its R covers every other pairing.
* `p_hat` = baseline composition used in the simulation.

```{r}
strata <- sel_long |>
  filter(condition == "k_fixed") |>
  group_by(model, instruction_type) |>
  group_modify(~ {
    m <- glm(y_uc ~ axis_cell + offset(log(K_u)), 
             family = poisson, 
             data = .x)
    # baseline composition, in cells order
    p <- tapply(.x$s_uc, .x$axis_cell, mean)[cells]
    tibble(phi_hat = sum(residuals(m, type = "pearson")^2) / df.residual(m),
           dev_ratio = deviance(m) / df.residual(m),
           spread = max(p) / min(p),
           target_cell = names(which.min(p)),
           target_actor = names(which.min(tapply(p, actor_of, sum))),
           p_hat = list(p))
  }) |>
  ungroup()
strata

# phi_p90 across the nine strata
quantile(strata$phi_hat, 0.90)
# narrowest spread (rr_cell should be about half of this or less)
min(strata$spread)
```

## 3. Simulation functions

Each simulated dataset is a two-group comparison: R runs drawn from a stratum's `p_hat` and R runs from the H1 composition, where the target group's share is multiplied by RR. Multinomial draws keep counts summing to K = 15 within a run, which reproduces the underdispersion in step 2. The test is a Poisson fit with robust SEs, one row per run, matching the independence GEE in Stage 2.

```{r}
# H1 composition: target group x rr, other cells rescaled so the shares sum to 1
make_p1 <- function(p0, hit, rr) {
  s0 <- sum(p0[hit])
  p0[hit] <- p0[hit] * rr
  p0[!hit] <- p0[!hit] * (1 - s0 * rr) / (1 - s0)
  p0
}

# p-value for one simulated dataset with r runs per group
p_value_once <- function(r, p0, p1, hit) {
  # draws are cells x runs, colSums of the target rows = target count per run
  y <- c(colSums(rmultinom(r, k_budget, p0)[hit, , drop = FALSE]),
         colSums(rmultinom(r, k_budget, p1)[hit, , drop = FALSE]))
  grp <- rep(0:1, each = r)
  # K is constant, so the log(K) offset is absorbed by the intercept
  m <- glm(y ~ grp, 
           family = poisson)
  V <- vcovHC(m, type = "HC0")
  2 * pnorm(-abs(coef(m)[["grp"]] / sqrt(V["grp", "grp"])))
}

# smallest R on the grid reaching the power target (NA = grid exhausted)
required_R <- function(p0, hit, rr) {
  p1 <- make_p1(p0, hit, rr)
  for (r in r_grid) {
    power <- mean(replicate(n_sim, p_value_once(r, p0, p1, hit)) < alpha)
    if (power >= power_t) return(r)
  }
  NA_integer_
}
```

## 4. Required R and decision

`quantile(type = 1)` returns an R that was actually simulated, so at least 80% of strata (8 of 9) reach the power target. The default interpolated quantile can land between grid values and cover fewer (e.g., 148 covers 7 of 9 on the actor tier). The hardest stratum is deliberately not covered by the 80th percentile criterion.

```{r}
set.seed(2026)
req_tbl <- strata |>
  rowwise() |>
  mutate(r_cell = required_R(p_hat, 
                             cells == target_cell, 
                             rr_cell),
         r_actor = required_R(p_hat, 
                              actor_of == target_actor, 
                              rr_actor)) |>
  ungroup()

# check that a stratum has not exhausts the grid would silently drop out of the quantile - widen r_grid instead
stopifnot(!anyNA(req_tbl$r_cell), 
          !anyNA(req_tbl$r_actor))

# 80th percentile across strata, larger of the two tiers
r_stage2 <- max(quantile(req_tbl$r_cell, 
                         cover_q, 
                         type = 1),
                quantile(req_tbl$r_actor, 
                         cover_q, 
                         type = 1))

req_tbl |>
  select(model, instruction_type, 
         target_cell, r_cell, 
         target_actor, r_actor)
r_stage2

# strata reaching the power target at r_stage2 (of 9)
req_tbl |>
  summarise(cell = sum(r_cell <= r_stage2),
            actor = sum(r_actor <= r_stage2))
```

## 5. Validation

Null calibration in the hardest stratum (smallest expected counts, where an asymptotic test is most likely to drift): with rr = 1 the rejection rate at `r_stage2` should sit near alpha, i.e. the binomial CI should cover 0.05.

```{r}
set.seed(2026)
# stratum with the largest r_cell
worst <- which.max(req_tbl$r_cell)
p0 <- req_tbl$p_hat[[worst]]
hit <- cells == req_tbl$target_cell[worst]

# p1 = p0 (no effect)
p_null <- replicate(n_sim, 
                    p_value_once(r_stage2, p0, 
                                 p0, hit))
binom.test(sum(p_null < alpha), 
           n_sim, 
           p = alpha)
```
